Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs

### 2014-04-11 paper/manuscript.tex
Working on the RNA-seq validation (2).

%%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
The CM based approach works well for most classes of ncRNA, but the
long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. Whilst some these regions may not reflect
functional RNA elements instead regulatory regions, enhancers or
insulators, their syntenic conservation still provides an indication
of lncRNA conservation \cite{diederichs2014four}.

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 predominantly non-model bird
  species for ncRNAs that are conserved across the avian species. This
  conservative set of annotations is expected to contain the core
  avian ncRNAs.}  We focus our report on the unusual results within
the avian lineages. These are either unexpectedly well-conserved
ncRNAs or unexpectedly poorly-conserved ncRNAs. The former are ncRNAs
that are not expected to be conserved between the birds and the other
vertebrates, particularly those ncRNAs whose function is not conserved
in birds.  The latter are apparent losses of ncRNAs expected to be
conserved; Here, we consider three categories of such ``loss'': First,
genuine gene losses in the avian lineage where ncRNAs well conserved
in other vertebrates are completely absent in birds.  Second,
``divergence'' where ncRNAs have undergone such significant sequence
and structural alternations that homology search tools can no longer
detect a relationship between other vertebrate exemplars and avian
varieties.  Third, ``missing'' ncRNAs that failed to be captured in
the available, largely fragmented, avian genomes.  We postulate that
the latter category is likely to be prevalent in comparative avian
genome studies given the distinctive organisation of the avian
genome. Namely, the avian karyotype is characterised by a large number
of chromosomes (average $2n \approx 80$) generally consisting of a few
larger ``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

>>>>

%%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
The CM based approach is appropriate for most classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. 
%Recent technological advances have led to
%dramatic speed and memory-usage enhancements for CM analyses
%\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. 
CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
they deal elegantly with the repeats that many lncRNAs
host. Consequently in the latest release of Rfam the lncRNA families
that were added were composed of local conserved (and possibly
structured elements) within lncRNAs, analogous to the ``domains''
housed within protein sequences \cite{Burge:2013}. Whilst some these
regions may not reflect functional RNA elements but instead regulatory
regions, enhancers or insulators, their syntenic conservation still
provides an indication of lncRNA conservation
\cite{diederichs2014four}.

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. 
{\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 predominantly non-model bird
  species for ncRNAs that are conserved across the avian species. This
  conservative set of annotations is expected to contain the core
  avian ncRNAs.}  We focus our report on the unusual results within
the avian lineages. These are either unexpectedly well-conserved
ncRNAs or unexpectedly poorly-conserved ncRNAs. The former are ncRNAs
that are not expected to be conserved between the birds and the other
vertebrates, particularly those ncRNAs whose function is not conserved
in birds.  The latter are apparent losses of ncRNAs expected to be
conserved; Here, we consider three categories of such ``loss'': First,
genuine gene losses in the avian lineage where ncRNAs well conserved
in other vertebrates are completely absent in birds.  Second,
``divergence'' where ncRNAs have undergone such significant sequence
and structural alternations that homology search tools can no longer
detect a relationship between other vertebrate exemplars and avian
varieties.  Third, ``missing'' ncRNAs that failed to be captured in
the available, largely fragmented, avian genomes.  We postulate that
the latter category is likely to be prevalent in comparative avian
genome studies given the distinctive organisation of the avian
genome. Namely, the avian karyotype is characterised by a large number
of chromosomes (average $2n \approx 80$) generally consisting of a few
larger ``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

The collection of ncRNA sequences is generally biased towards model
organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
that using data from well studied lineages such as mammals can result
in quality annotations of sister taxa such as aves.
